---
title: "p8105_mtp_yx2954"
author: "Yiran Xu"
date: "2024-10-18"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
library(knitr)
library(rvest)
library(tidyverse)
library(scales)
```

# Problem 1

Goal: take a look on the trend of home value across US and rent within NYC from three raw dataset:

*   zhvi_df: ZIP code level house price in each month of 2023
*   zori_df: ZIP code level rent in each month between January 2015 and August 2024 
*   zip_df: neighborhood to which each ZIP code attaches

First tidy NYC rental data to make all prices in a single column, the clean the column names.
```{r}
zori_df = read_csv("data/Zillow/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") |>
  pivot_longer(cols = "2015-01-31":"2024-08-31",
               names_to = "date",
               values_to = "rent") |>
  mutate(date = as.Date(date), 
         zip = RegionName)|>
  janitor::clean_names() 
```

Create a new column to store the corresponding borough name according to the county name, and clean variable name. 
```{r}
url = "https://p8105.com/data/zip_codes.html"
zip_list = read_html(url) |>
  html_table() 

zip_df = 
  zip_list[[1]] |>
  mutate(borough = case_when(
    County == "New York" ~ "Manhattan",
    County == "Kings" ~ "Brooklyn",
    County == "Queens" ~ "Queens",
    County == "Bronx" ~ "Bronx",
    County == "Richmond" ~ "Staten Island")) |>
  janitor::clean_names() 
```

Duplicate Zip codes are identified
```{r}
duplicate_zip = 
  zip_df |>
  filter(duplicated(zip_code))
```

```{r}
duplicate1 = 
  zip_df |>
  filter(zip_code == "10463")

duplicate1
```

```{r}
duplicate2 = 
  zip_df |>
  filter(zip_code == "11201")

duplicate2
```
We know Kingsbridge and Northwest Brooklyn are not in Manhattan. Remove the error row for quality control
```{r}
zip_final_df = 
  zip_df |> 
  filter(!(borough == "Manhattan" & zip_code == "11201")) |>
  filter(!(borough == "Manhattan" & zip_code == "10463")) 
```

Then merge two date set and only keep neccessary columns
```{r}
zori_final_df = left_join(zori_df, zip_final_df, by = c("region_name" = "zip_code")) |>
  select(-size_rank, -region_id, -region_name:-county_name, -county:-file_date) 
```


```{r}
uni_zip = zori_final_df |>
  pull(zip) |>
  n_distinct()

uni_nbh = zori_final_df |>
  pull(neighborhood) |>
  n_distinct()
```
After tidy and clean the dataset, there are **`r nrow(zori_final_df)`** total observations exist. **`r uni_zip`** unique ZIP codes. **`r uni_nbh`** unique neighborhoods


First put all housing prices across time in a single column, then clean column name into lower cases with underscore marks, and keep minimal number of variables needed.

```{r}
zhvi_df = read_csv("data/Zillow/Zip_zhvi_uc_sfrcondo_tier_0.33_0.67_sm_sa_month_2023.csv") |>
  pivot_longer(
    cols = starts_with("2023"),
    names_to = "date",
    values_to = "price"
  ) |>
  janitor::clean_names() |>
  mutate(zip = region_name,
           date = as.Date(date)) |>
  select(-size_rank, -region_id, -region_name, -region_type, -state_name, -metro: -county_name) 

head(zhvi_df)
```

# Problem 2

Not all ZIP codes has 116 observations. Because some house were not being rent in some period and some may still under construction

```{r}
zori_nzip = 
  zori_df |>
  pull(zip) |>
  n_distinct()

zip_nzip = 
  zip_df |>
  pull(zip_code) |>
  n_distinct()
```

There are **`r zori_nzip`** different ZIP in NYC Rental data and **`r zip_nzip`** different ZIP in zip code data. This **`r zip_nzip - zori_nzip`** discrepancy could be explained by some non-residental ZIP code included in the zip code data and areas where no house were for rent or the rental was not on Zillow platform.


```{r}
avg_rent = zori_final_df |>
  group_by(borough, year(date)) |>
  summarize(mean_price = mean(rent, na.rm = TRUE)) |>
  pivot_wider(
    names_from = borough,
    values_from = mean_price
  )

kable(avg_rent, caption = "Average Rent by Borough and Year", col.names = c("year", "Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island"))
```

The average rental price increases with time across all boroughs. A sudden decrease in the price was observed from 2019 to 2020 in Brooklyn, Manhattan, and Queens. Manhattan has the highest rental price, followed by Brooklyn, Queens, Staten Island and Bronx.

```{r}
rent_drop_table = zori_final_df |>
  filter(date %in% c("2020-01-31", "2021-01-31")) |>
  mutate(date = year(date)) |>
  pivot_wider(names_from = date, values_from = rent, names_prefix = "year_") |>
  mutate(rent_drop = year_2020 - year_2021) |>
  group_by(borough) |>
  filter(rent_drop == max(rent_drop, na.rm = TRUE)) |>
  select(borough, zip, neighborhood, rent_drop)

kable(rent_drop_table, caption = "Max Rent Drop from 2020-2021 by Borough", col.names = c("borough", "zip", "neighborhood", "rent_drop"))
```

The rental price in Bronx has almost no changes, while that in other 3 boroughs except Staten Island has dropped from Jan 2020 to Jan 2021. The drop in Manhattan (**`r rent_drop_table[4, 4]`**) was much higher than Queens (**`r rent_drop_table[1, 4]`**) and Brooklyn (**`r rent_drop_table[2, 4]`**). No data was available in Staten Island.

# Problem 3

```{r zori_year_p, fig.path='results/', fig.ext='png'}
zori_year_p = 
  zori_final_df |>
  mutate(year = year(date)) |>
  group_by(borough, zip, year) |>
  summarise(avg_price = mean(rent, na.rm = TRUE)) |>
  ggplot(aes(x = as.factor(year), y = avg_price)) +
  geom_boxplot() +  
  facet_grid(~ borough, scales = "free_y") +
  labs(title = "NYC Rental Prices Distribution by Year and Borough",
       x = "Year", y = "Rental Price") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(labels = comma)

zori_year_p
```

Rental prices increased in the past 10 years. While being higher and dispersedly distributed in Brooklyn and Manhattan, the rental prices are more evenly distributed and lower in the other 3 boroughs. Upper outliers are observed in Manhattan and Queens, meaning  some luxury apartment available. The Staten Island is lack of data prior to 2020.

```{r}
zhvi_state = zhvi_df |>
  mutate(year = year(date),
         month = month(date)) |>
  group_by(zip, state) |>
  summarise(avg_price = mean(price, na.rm = TRUE)) |>
  arrange(desc(avg_price)) 
```

```{r zhvi_state_p, fig.path='results/', fig.ext='png'}
zhvi_state_p = 
  zhvi_state |>
  mutate(avg_price = as.numeric(avg_price)) |>
  group_by(state) |>
  mutate(median_price = median(avg_price, na.rm = TRUE)) |>
  ungroup() |>
  ggplot(aes(x = fct_reorder(as.factor(state), median_price), y = avg_price)) +
  geom_boxplot() +  
  labs(title = "Average Housing Prices in 2023 Distribution by State",
       x = "State", y = "House Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(
    trans = "sqrt", 
    position = "right",
    labels = comma)

zhvi_state_p
```

```{r}
median_price_df =
  zhvi_state |> 
  group_by(state) |>
  summarise(median_price = median(avg_price, na.rm = TRUE)) |>
  arrange(desc(median_price))
```

The plot above shows the average ZIP code level housing price in 2023 in different state. The five states with the highest median house prices are **`r median_price_df[1, 1]`** (**`r median_price_df[1, 2]`**), **`r median_price_df[2, 1]`** (**`r median_price_df[2, 2]`**), **`r median_price_df[3, 1]`** (**`r median_price_df[3, 2]`**), **`r median_price_df[4, 1]`** (**`r median_price_df[4, 2]`**), **`r median_price_df[5, 1]`** (**`r median_price_df[5, 2]`**). Most outlier are found above the boxes, indicating some luxury houses available. The highest housing price was in **`r zhvi_state[1, 2]`** with a price of **`r zhvi_state[1, 3]`**

```{r}
zhvi_NYC_df =
  zhvi_df |>
  filter(city == "New York") |>
  mutate(year = year(date)) |>
  group_by(zip) |> 
  summarize(avg_price = mean(price)) |>
  mutate(zip = as.numeric(zip))
```

```{r}
zori_month_df = 
  zori_final_df |>
  mutate(year = year(date)) |>
  filter(year == "2023") |>
  group_by(zip) |>
  summarise(avg_rent = mean(rent))
```

```{r rent_price_p, fig.path='results/', fig.ext='png'}
rent_price_p = 
  inner_join(zori_month_df, zhvi_NYC_df, by = "zip") |>
  ggplot(aes(x = avg_rent, y = avg_price)) +
  scale_y_continuous(labels = comma, limits = c(0, 4000000)) +
  scale_x_continuous(labels = comma, limits = c(2000, 5000)) +
  labs(title = "ZIP-code-specific housing prices against ZIP-code-specific rental prices",
       x = "Rental Price", y = "Housing Price") +
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) 

rent_price_p
```

Most housing prices are distributed from 500,000 to 1,500,000; most rental prices range from 2,000 to 4,000. Housing prices and rental prices are positively correlated.


```{r}
zip_diff = anti_join(zhvi_NYC_df, zori_month_df, by = "zip") |>
  nrow()

num_SI = 
  zori_final_df |>
  filter(borough == "Staten Island") |>
  pull(zip) |>
  n_distinct()
```

Limitations: 

*   Rental price dataset has **`r zip_diff`** fewer ZIP code available compared to housing price dataset
*   Only **`r num_SI`** ZIP code in Staten Island is presented. 
*   Only housing price in 2023 are available 
*   Incomplete time coverage for each ZIP code
